---
title: "Reproducible Research Assignment 1"
output: html_document
---

The first part is loading and preprocessing the data:

The following code loads and processes the data i.e. (activity.csv) given for the assignment:

```{r, echo = TRUE}
# Load the raw activity data
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)

#Transform the date attribute to an actual date format
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Compute the weekdays from the date attribute
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the day type (weekend or weekday)
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                     activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

# Clear the workspace
rm(activity_raw)
```

The following code displays the first few rows of the activity data frame:

```{r, echo = TRUE}
head(activity)
```

The second part of the assignmnet is to calculate the mean total number of steps taken per day.
For this part of the assignment, the missing values are ignored in the dataset.

```{r, echo = TRUE}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)

# Rename the attributes
names(sum_data) <- c("date", "total")
```

The first few rows of the sum_data data frame are displayed as under:

```{r, echo = TRUE}
head(sum_data)
```
The first part of this question is to make a histogram of the total number of steps taken each day.
The histogram is given by the following lines of code:

```{r, echo = TRUE}
# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 20), 
     main="Histogram of the total number of steps taken each day\n(NA removed)")
```
The second question was to calculate and report the mean and median total number of steps taken per day.
The mean and median are computed withe the following code:

```{r, echo = TRUE}
mean(sum_data$total)
median(sum_data$total)
```

The third part of the assignment is to know what is the average daily activity pattern?
The first thing to do in this part is to make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):

```{r, echo = TRUE}
# Clear the workspace
rm(sum_data)

# Compute the means of steps accross all days for each interval
mean_data <- aggregate(activity$steps, 
                       by=list(activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)

# Rename the attributes
names(mean_data) <- c("interval", "mean")
```

The first few rows of the mean_data data frame are displayed as under:

```{r, echo = TRUE}
head(mean_data)
```

The time serie plot is created by the following lines of code:

```{r, echo = TRUE}
# Compute the time series plot
plot(mean_data$interval, 
     mean_data$mean, 
     type="l", 
     col="blue", 
     lwd=2, 
     xlab="Interval [minutes]", 
     ylab="Average number of steps", 
     main="Time-series of the average number of steps per intervals\n(NA removed)")
```

The second question of this part is to ascertain which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 

The following code will give the result.

```{r, echo = TRUE}
# We find the position of the maximum mean
max_pos <- which(mean_data$mean == max(mean_data$mean))

# We lookup the value of interval at this position
max_interval <- mean_data[max_pos, 1]
print(max_interval)
```

The fourth part of the assignment is Inputing the missing values
as there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries.

The first question of this part is to calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA???s) which can be accomplished by the following code: 

```{r, echo = TRUE}
# Clear the workspace
rm(max_interval)

# We use the trick that a TRUE boolean value is equivalent to 1 and a FALSE to 0.
NA_count <- sum(is.na(activity$steps))
print(NA_count)
```
 The second aim is to devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. 
 
For accomplishing tht I used the strategy to remplace each NA value by the mean of the steps attribute as coded below.

```{r, echo = TRUE}
# Clear the workspace
rm(NA_count)

# Find the NA positions
na_pos <- which(is.na(activity$steps))

# Create a vector of means
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(na_pos))
```


The third question of this part is to create a new dataset that is equal to the original dataset but with the missing data filled in.

The same is accomplished by the following line of code.

```{r, echo = TRUE}
# Replace the NAs by the means
activity[na_pos, "steps"] <- mean_vec

# Clear the workspace
rm(mean_vec, na_pos)
```

We display the first few rows of the new activity data frame:

```{r, echo = TRUE}
head(activity)
```

The fourth question of this part is to make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Also we need to identify if these values differ from the estimates from the first part of the assignment? and also what is the impact of imputing missing data on the estimates of the total daily number of steps?
 The Histogram is plotted by the following chunk of code:
```{r, echo = TRUE}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
```

The mean and median are computed like:

```{r, echo = TRUE}
mean(sum_data$total)
median(sum_data$total)
```



The last part of the assignment is to identify if there are any  differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. I use the dataset with the filled-in missing values for this part.

The first question of this part is to create a new factor variable with the following code in the dataset with two levels - ???weekdays??? and ???weekend??? indicating whether a given date is a weekday or weekend day.

```{r,echo = TRUE}
# The new factor variable "daytype" was already in the activity data frame so I siply display the head of dataset.
head(activity)
```

The second question of this part is to make a panel plot containing a time series plot (i.e. type = "l") of the 5- minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo = TRUE}
# Clear the workspace
rm(sum_data)

# Load the lattice graphical library
library(lattice)

# Compute the average number of steps taken, averaged across all daytype variable
mean_data <- aggregate(activity$steps, 
                       by=list(activity$daytype, 
                               activity$weekday, activity$interval), mean)

# Rename the attributes
names(mean_data) <- c("daytype", "weekday", "interval", "mean")
```

The first few rows of the mean_data data frame are displayed as under:

```{r, echo = TRUE}
head(mean_data)
```

The time series plot is plotted as under:

```{r, echo = TRUE}
# Compute the time serie plot
xyplot(mean ~ interval | daytype, mean_data, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))
# Clear the workspace
rm(mean_data)
```


